Hazelnut Live: Type-Theoretic Foundations for Running Incomplete Programs

### 2017-05-28 calculus.tex
added figure labels and copied over syntax and type consistency defns from popl paper
macro for declarative H-expressions
draft of dynamics for nums, arrow, and sums
fix typos; add is-final to premises of i-add_1 and i-add_2
decorations on holes in dynamics
substitution
dunno
grammar
syntax figure caption
updated type consistency and matched types judgements with products and sums
commented out various complexities
algorithmic expansion rules
type assignment rules

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\todo{exposition}
\todo{add sums and products}

\begin{figure}[h]
...
\caption{Syntax of H-expressions and H-types.}
\end{figure}

\begin{figure}[h]
\judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
...

\judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
...
\caption{Type consistency and matched types.}
\end{figure}

\begin{figure}[h]
\judgbox{\expandSyn{\hGamma}{\hexp}{M}{\htau}{\Delta}}{ }
...

\judgbox{\expandAna{\hGamma}{\hexp}{M}{\htau}{\Delta}}{ }
...
\caption{Bidirectionally Typed Expansion}
\end{figure}

\begin{figure}[h!]
\judgbox{\hasType{\hGamma}{M}{\htau}{\Delta}}{$M$ has type $\htau$ with holes $\Delta$}
...
\caption{Declarative Statics}
\end{figure}

\begin{figure}[h!]
\judgbox{\isValue{M}}{$M$ is a value}
...

\judgbox{\isIndet{M}}{$M$ is indeterminate}
...

\judgbox{\stepsTo{M}{N}}{$M$ steps to $N$}
...
\caption{Structural Dynamics}
\end{figure}

>>>>

\clearpage
\newcommand{\calculusSec}{Hazelnut Live, Formally}
\section{\protect\calculusSec}
\label{sec:calculus}
\todo{exposition}

\begin{figure}[h]
$\arraycolsep=4pt\begin{array}{rllllll}
\mathsf{HTyp} & \htau & ::= &
  b ~\vert~
  \tarr{\htau}{\htau} ~\vert~
  % \tprod{\htau}{\htau} ~\vert~
  % \tsum{\htau}{\htau} ~\vert~
  \tehole\\
\mathsf{HExp} & \hexp & ::= &
  c ~\vert~ 
  x ~\vert~
  \halam{x}{\htau}{\hexp} ~\vert~
  \hap{\hexp}{\hexp} ~\vert~
  % \hpair{\hexp}{\hexp} ~\vert~
  % \hprj{i}{\hexp} ~\vert~
  % \hinj{i}{\hexp} ~\vert~
  % \hcase{\hexp}{x}{\hexp}{x}{\hexp} ~\vert~
  % \hadd{\hexp}{\hexp} ~\vert~
  \hehole{u} ~\vert~
  \hhole{\hexp}{u} ~\vert~
  {\hlam{x}{\hexp}} ~\vert~
  \hexp : \htau\\
 \mathsf{DHExp} & \dexp  & ::= &
  c ~\vert~
  x ~\vert~
  {\halam{x}{\htau}{\dexp}} ~\vert~
  \hap{\dexp}{\dexp} ~\vert~
  % \hpair{\dexp}{\dexp} ~\vert~
  % \hprj{i}{\dexp} ~\vert~
  % \hinj{i}{\dexp} ~\vert~
  % \hcase{\dexp}{x}{\dexp}{x}{\dexp} ~\vert~
  % \hadd{\dexp}{\dexp} ~\vert~
  \dehole{\mvar}{\subst} ~\vert~
  \dhole{\dexp}{\mvar}{\subst} ~\vert~
  \dcast{\dexp}{\htau}\\
\end{array}$
\caption{Syntax of H-types, H-expressions and dynamic H-expressions. We write $x$ to range over variables, $u$ over hole names and $\sigma$ over finite substitutions.}
\label{fig:HTyp}
\label{fig:HExp}
\end{figure}

\begin{figure}[h]
\judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
\begin{mathpar}
\inferrule{ }{
  \tconsistent{\tehole}{\htau}
}

\inferrule{ }{
  \tconsistent{\htau}{\tehole}
}

\inferrule{ }{
  \tconsistent{\htau}{\htau}
}

\inferrule{
  \tconsistent{\htau_1}{\htau_1'}\\
  \tconsistent{\htau_2}{\htau_2'}
}{
  \tconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
}
%
% \inferrule{
%   \tconsistent{\htau_1}{\htau_1'}\\
%   \tconsistent{\htau_2}{\htau_2'}
% }{
%   \tconsistent{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1'}{\htau_2'}}
% }
%
% \inferrule{
%   \tconsistent{\htau_1}{\htau_1'}\\
%   \tconsistent{\htau_2}{\htau_2'}
% }{
%   \tconsistent{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1'}{\htau_2'}}
% }
\end{mathpar}

\judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{mathpar}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}

\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{mathpar}

% \judgbox{\prodmatch{\htau}{\tprod{\htau_1}{\htau_2}}}{$\htau$ has matched product type $\tprod{\htau_1}{\htau_2}$}
% \begin{mathpar}
% \inferrule{ }{
%   \prodmatch{\tehole}{\tprod{\tehole}{\tehole}}
% }

% \inferrule{ }{
%   \prodmatch{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1}{\htau_2}}
% }
% \end{mathpar}

% \judgbox{\summatch{\htau}{\tsum{\htau_1}{\htau_2}}}{$\htau$ has matched sum type $\tsum{\htau_1}{\htau_2}$}
% \begin{mathpar}
% \inferrule{ }{
%   \summatch{\tehole}{\tsum{\tehole}{\tehole}}
% }

% \inferrule{ }{
%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
% }
% \end{mathpar}
\caption{Type Consistency and Matched Types}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

\begin{figure}[h]
\judgbox{\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}}{ }
\begin{mathpar}
\inferrule[expand-constant]{ }{
  \expandSyn{\hGamma}{c}{b}{c}{\emptyset}
}

\inferrule[expand-var]{ }{
  \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
}

\inferrule[expand-alam]{
  \expandSyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\Delta}
}{
  \expandSyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\Delta}
}

\inferrule[expand-ap-1]{
  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}\\
  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\dexp_1}{\tarr{\htau_2}{\tehole}})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
}

\inferrule[expand-ap-2]{
  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau'_2}{\Delta_2}\\
  \htau_2 \neq \htau'_2
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\dexp_2}{\htau_2}}}{\Dunion{\Delta_1}{\Delta_2}}
}

\inferrule[expand-ap-3]{
  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau_2}{\Delta_2}
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
}\\
%
% \inferrule[expand-pair]{
%   \expandSyn{\hGamma}{\hexp_1}{\htau_1}{\dexp_1}{\Delta_1}\\
%   \expandSyn{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\Delta_2}
% }{
%   \expandSyn{\hGamma}{\hpair{\hexp_1}{\hexp_2}}{\tprod{\htau_1}{\htau_2}}{\hpair{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
% }
%
% \inferrule[expand-prj]{
%   a
% }{
%   b
% }
%
% (inj)

%
% \inferrule[expand-plus]{ }{
%   \expandSyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}{\hadd{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
% }
%
\inferrule[expand-syn-empty-hole]{ }{
  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}}{\Dbinding{u}{\hGamma}{\tehole}}
}

\inferrule[expand-syn-non-empty-hole]{
  \expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}
}{
  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
}\\
%
\inferrule[expand-asc-1]{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}\\
  \htau \neq \htau'
}{
  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\dexp}{\htau}}{\Delta}
}

\inferrule[expand-asc-2]{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}
}{
  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dexp}{\Delta}
}
\end{mathpar}

\judgbox{\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}}{ }
\begin{mathpar}
\inferrule[expand-lam]{
  \expandAna{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\htau'_2}{\Delta}
}{
  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2'}}{\Delta}
}

\inferrule[expand-subsume]{
  \hexp \neq \hehole{u}\\
  \hexp \neq \hhole{\hexp'}{u}\\\\
  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
  \tconsistent{\htau}{\htau'}
}{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}
}

\inferrule[expand-ana-empty-hole]{ }{
  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
}

\inferrule[expand-ana-non-empty-hole]{ }{
  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
}
\end{mathpar}
\caption{Algorithmic Expansion}
\label{fig:expandSyn}
\label{fig:expandAna}
\end{figure}

\begin{figure}[h!]
\judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
\begin{mathpar}
\inferrule[ty-assign-const]{ }{
  \hasType{\Delta}{\hGamma}{c}{b}
}

\inferrule[ty-assign-var]{ }{
	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
}

\inferrule[ty-assign-lam]{
  \hasType{\Delta}{\hGamma, x : \htau_1}{\dexp}{\htau_2}
}{
  \hasType{\Delta}{\hGamma}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2}}
}

\inferrule[ty-assign-ap]{
  \hasType{\Delta}{\hGamma}{\dexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
  \hasType{\Delta}{\hGamma}{\dexp_2}{\htau_2}
}{
  \hasType{\Delta}{\hGamma}{\hap{\dexp_1}{\dexp_2}}{\htau}
}

\inferrule[ty-assign-empty-hole]{
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}}{\htau}
}

\inferrule[ty-assign-non-empty-hole]{
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}}{\htau}
}

\inferrule[ty-assign-cast]{
  \hasType{\Delta}{\Gamma}{\dexp}{\htau'}\\
  \tconsistent{\htau}{\htau'}
}{
  \hasType{\Delta}{\hGamma}{\dcast{\dexp}{\htau}}{\htau}
}
\end{mathpar}
\caption{Type Assignment for Dynamic H-Expressions}
\label{fig:hasType}
\end{figure}

\begin{definition}[Substitution Type Assignment]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
\end{definition}

\begin{figure}[h!]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcl}
[\dexp_1 / x] \dehole{\mvar}{\subst}
&=&
\dehole{\mvar}{[\dexp_1 / x] \cdot \subst}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
&=& 
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
\\
{[\dexp_1 / x] \dlam{y}{\dexp_2}}
&=&
{\dlam{y}{[\dexp_1 / x] \dexp_2}}
\\
{[\dexp_1 / x] \dapP{e_2}{e_3}}
&=&
{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\\
{[\dexp_1 / x] \dinj{i}{e_2}}
&=&
{\dinj{i}{[\dexp_1 / x] e_2}}
\\
{[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
&=&
{\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
\\
{[\dexp_1 / x] \dnum{n}}
&=&
{\dnum{n}}
\\
{[\dexp_1 / x] \dadd{e_2}{e_3}}
&=&
{\dadd{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\end{array}
\]
\end{figure}

\begin{figure}
\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
\begin{mathpar}
\Infer{v-lam}
{ }{\isValue{\hlam{x}{\dexp}}}
\and
\Infer{v-num}
{ }{\isValue{\hnum{n}}}
\and
\Infer{v-inj}
{\isValue{\dexp}}{\isValue{\hinj{i}{\dexp}}}
\end{mathpar}

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\Infer{i-ehole}
{ }
{\isIndet{\dehole{\mvar}{\subst}}}
\and
\Infer{i-hole}
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}
\\
\Infer{i-add$_1$}
{\isIndet{\dexp_1} \\ \isFinal{\dexp_2}}
{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{i-add$_2$}
{\isFinal{\dexp_1} \\ \isIndet{\dexp_2}}
{\isIndet{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{i-app}
{\isIndet{\dexp_1}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}
\and
\Infer{i-case}
{\isIndet{\dexp}}
{\isIndet{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
\end{mathpar}

\judgbox{\isErr{\dexp}}{$\dexp$ is error}
\begin{mathpar}
\Infer{e-cast}
% XXX use macro type assignment judgement here
{ \vdash e : \htau_2 \\ \tinconsistent{\htau_1}{\htau_2}}
{\isErr{\dcast{\dexp}{\htau_1}}}
\and
\Infer{e-add$_1$}
{\isErr{\dexp_1}}
{\isErr{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{e-add$_2$}
{\isErr{\dexp_2}}
{\isErr{\dadd{\dexp_1}{\dexp_2}}}
\and
\Infer{e-app}
{\isErr{\dexp_1}}
{\isErr{\dap{\dexp_1}{\dexp_2}}}
\and
\Infer{e-case}
{\isErr{\dexp}}
{\isErr{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}}
\end{mathpar}

\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
\begin{mathpar}
\Infer{f-val}
{\isValue{\dexp}}{\isFinal{\dexp}}
\and
\Infer{f-indet}
{\isIndet{\dexp}}{\isFinal{\dexp}}
\and
\Infer{f-err}
{\isErr{\dexp}}{\isFinal{\dexp}}
\end{mathpar}

\judgbox{\stepsTo{\dexp_1}{\dexp_2}}{$\dexp_1$ steps to $\dexp_2$}
\begin{mathpar}
\Infer{s-hole}
{\stepsTo{\dexp_1}{\dexp_2} }
{\stepsTo{\dhole{\dexp_1}{\mvar}{\subst}}{\dhole{\dexp_2}{\mvar}{\subst}}}
\and
\Infer{s-cast}
% XXX use macro type assignment judgement here
{ \vdash e : \htau_2 \\ \tconsistent{\htau_1}{\htau_2}}
{\stepsTo{\dcast{\dexp}{\htau_1}}{\dexp}}
\\
\Infer{s-add$_1$}
{\stepsTo{\dexp_1}{\dexp_1'}}
{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1'}{\dexp_2}}}
\and
\Infer{s-add$_2$}
{\isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
{\stepsTo{\dadd{\dexp_1}{\dexp_2}}{\dadd{\dexp_1}{\dexp_2'}}}
\and
\Infer{s-add$_3$}
{ n_1 + n_2 = n_3 }
{\stepsTo{\dadd{\dnum{n_1}}{\dnum{n_2}}}{\dnum{n_3}}}
%\and
\\
\Infer{s-app$_1$}
{\stepsTo{\dexp_1}{\dexp_1'}}
{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1'}{\dexp_2}}}
\and
\Infer{s-app$_2$}
{ \isFinal{\dexp_1} \\ \stepsTo{\dexp_2}{\dexp_2'}}
{\stepsTo{\dap{\dexp_1}{\dexp_2}}{\dap{\dexp_1}{\dexp_2'}}}
\and
\Infer{s-app$_\beta$}
{ \isFinal{\dexp_2} }
{\stepsTo{\dapP{\dlam{x}{\dexp_1}}{\dexp_2}}{ [\dexp_2/x]\dexp_1 }}
%\and
\\
\Infer{s-case$_1$}
{\stepsTo{\dexp}{\dexp'}}
{\stepsTo{\dcase{\dexp}{x}{\dexp_x}{y}{\dexp_y}}{\dcase{\dexp'}{x}{\dexp_x}{y}{\dexp_y}}}
\and
\Infer{s-case$_2$}
{ }
{\stepsTo{\dcase{\dinj{1}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/x]\dexp_x }}
\and
\Infer{s-case$_3$}
{ }
{\stepsTo{\dcase{\dinj{2}{\dexp}}{x}{\dexp_x}{y}{\dexp_y}}{ [\dexp/y]\dexp_y }}
\end{mathpar}
\caption{Structural Dynamics}
\label{fig:isValue}
\label{fig:isIndet}
\label{fig:stepsTo}
\end{figure}
